Investigating Neolithic pottery use from Guijiabao in the Yanyuan Basin in southwest China using organic residue analysis

### 2023-04-15 analysis/paper/paper.Rmd
minor edits before submission

---
title: "Investigating Neolithic pottery use from Guijiabao in the Yanyuan Basin in southwest China using organic residue analysis"
author:
  - Li-Ying Wang:
      email: liyingw@asihp.net
      institute: [AcS]
      correspondence: true
  - Kuei-chen Lin:
      email: kueichenlin@asihp.net
      institute: [AcS]
      correspondence: false
  - Zhiqing Zhou:
      email: zhouzhiQ2000@163.com
      institute: [InsC]
      correspondence: false
institute:
  - AcS: Institute of History and Philology, Academia Sinica 
  - InsC: Chengdu Institute of Cultural Relics and Archaeology
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
    The spread of new crops or terrestial animals across areas in southwest China during the Neolithic period is one of the important archaeological issues to understand the process of adoption of new culinary practices in those areas. It is widely accepted that both millet and rice were introduced into southwest China in 4700 BP at the latest. However, to what extent new crops were used by local peoples at an earlier time remains unclear. The excavation of Guijiabao in the Yanyuan basin in southwest China uncovered abundant materials with well-preserved stratigraphic records and has provided a unique opportunity to understand the local adoption of new crops. To further understand pottery use and food practices at Guijiabao, we investigated organic residues preserved in potsherds using lipid biomarker identification and compounds-specific stable isotope analysis. The results show signals of terrestrial animals in most potsherds and a C3 plant source in one potshard, during the Neolithic period. Our case study indicates that hunting might have been the primary means for obtaining subsistence, supplemented by rice/millet farming. This study also demonstrates an approach using organic residue analysis to explore pottery use and culinary practices during the Neolithic period in southwest China. A more systematic sampling of different types of potsherds and temporal phases with larger samples is needed to provide a clearer picture of ancient diets and subsistence.

@@

# Introduction
<!-- intro to the topic and research aim-->
Movements of people between geographically distinct areas are usually accompanied by the exchange of goods, technology, or ideas that could have cultural or economic impacts on any people groups that are involved in the interactions that result from these movements [@Burmeister2000]. The Hengduan Mountain ranges and valleys in southwestern China are geographical areas that connect the Qinghai-Tibetan Plateau and the Yunnan–Guizhou Plateau and by which movement to these plateaus must occur. They play an important role in the cultural exchanges and spread of crops into southwestern China and even further to mainland Southeast Asia [@Ma2021; @Lin2022; @Hao2022; @Tong1986]. Despite a long-standing debate on the routes of rice/millet dispersal from northwest China and the middle Yangtze valley to southwestern China [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming was practiced in southwest Sichuan around 4700 BP and in western Yunnan before 4600 BP [@Jiang2016; @YunnanMuseum1981; @Hao2022; @Zhang2010; @Dal2022; @Chen2015]. The Hengduan Mountain regions in the Neolithic period are therefore essential for understanding the processes of agricultural expansion. Recent archaeological evidence from the Yanyuan Basin in the middle Hengduan Mountains further reveals that millet and rice were present at roughly the same period as the grains found in Sichuan [@Huan2022; @Hao2022]. However, to what extent the two crops were incorporated into the local diet is still not fully understood.

To explore whether new crops or terrestrial animals became important food sources during the Neolithic period in the Yanyuan Basin, we investigated food residues within ceramic vessels from Guijiabao, a long-term occupation settlement with complete temporal sequences from the Neolithic period to the historical period in Yanyuan that have been excavated so far [@Guijiabao2017a; @Guijiabao2020a]. The new evidence at Guijiabao reveals that millet and rice were both present there in the Neolithic period around 5000 BP, earlier than previously thought [@Guijiabao2016], and millet could have already been part of the local diet [@Lin2022]. Although some of these analyses suggest the presence of mixed millet/rice farming, we still lack direct evidence of culinary practices during this period. To understand the extent of consuming new crops or terrestrial animals at Guijiabao, we ask the questions: What were the culinary practices that can be deduced by the pottery? Can we identify the new crops or terrestrial animals that were incorporated into the local food diets during the Neolithic period? We answer these questions by examining potsherds using organic residue analysis, which allows us to directly explore food cooked or processed in ceramic vessels.

The recent growing application of organic residue analysis demonstrates its usefulness in understanding ancient pottery use [@Kimpe2004; @Mayyas2022], human subsistence strategies, or culinary practices across a region over time [@Kwak2020; @Craig2011; @Suryanarayan2021; @Zhang2022; @Junno2020]. Examining organic residues, particularly lipids, enables archaeologists to investigate ancient food practices through lipid distributions and specific lipid compounds as biomarkers for tracing food origins [@Copley2005; @Whelton2021]. This is possible due to the protection of clay matrices from microbiological degradation and water leaching, which allows lipid molecules to accumulate and be well preserved in the porous fabric of pottery for over a thousand years [@Evershed1993; @Evershed2008; @Roffet2017]. In this study, we detected and identified lipid residues absorbed into the ceramic matrix that can reveal the original contents cooked or stored in a vessel during its lifetime of use [@Dudd1999; @Miller2020]. Through our case study, We attempt to investigate culinary practices during a period of frequent cultural interactions that accompanied agricultural expansion .

@@

<!--Yanyuan area and Guijiabao site-->
We focus on the Guijiabao site located in the Yanyuan Basin, which is a pull-apart basin in the middle eastern part of the Hengduan Mountain ranges (Figure \@ref(fig:SW-china-map)). The Guijiabao site allows us to understand the early adoption of agriculturebecause of its uniqueness and rarity: it contains complete cultural components from the Neolithic period to the historical period in that region [@Guijiabao2017a; @Guijiabao2017b]. A survey in 2015 and subsequent excavations from 2016 to 2019 revealed abundant archaeological materials and features, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guijiabao was a settlement site [@Hao2022]. The chronology of the Guijiabao site can be divided into three phases, the Neolithic period (5000-3700 BP), the Bronze Age (3200-2700 BP), and the late North Song to early Ming dynasties (1200-500 BP) based on 19 radiocarbon dates from plant remains and a human bone. For the Neolithic period, it can be further divided into two subphases, 5000-4500 BP and 4500-3700 BP, according to the content of artifact assemblages and ceramic forms and styles that were identified [@Hao2022].

@@

Although the recent findings at the Guijiabao and Gaoshan sites support mixed agriculture as the primary subsistence strategy, it is still unclear how much millet or rice was included in the local diets during the Neolithic period. The macro-botanical and phytolith analyses of the Guijiabao site show that millet was the major crop, similar to the findings at the Yingpanshan site in Sichuan [@Zhao2011]. However, plant remains from many sites in southwest China reveal a predominance of rice, such as at the Gaoshan and Baodun sites in Sichuan [@JiangGaoshan2017] and at Henglanshan and Baiyangcun sites in Yunnan [@Jiang2016; @Henglanshan2014]. This indicates that there were variations in agricultural practices in southwest China. To better understand the extent of the adoption of new crops, organic residue analysis of ceramics is a useful approach since we can directly study food residues related to culinary practices, including food preparation, processing, and cooking. This provides another way to examine the expansion of agriculture in the Yanyuan area.

@@

For sample processing, the surface of each potsherd was removed to prevent external contamination from environmental factors and handling during excavation or post-excavation analysis. Approximately 3.5 grams of powdered samples were extracted with a mixture of dichloromethane/methanol solution (9:1 v/v) using accelerated solvent extraction (ASE Dionex 350; at 100 °C). We conducted saponification by adding 10 mL of 1M KOH in methanol and 1 mL water to the extracts and heating them in an oven at 70 °C overnight. Saponified extracts were acidified with 6M hydrochloric acid (HCl), and the acid fraction was extracted using hexane ($3\times10$ mL). A 5% aliquot of each sample was derivatized using 20 uL N,O-bis(trimethylsilyl)trifluoroacetamide (BSTFA) with 1% v/v trimethylchlorosilane (TMCS), and heated at 70 °C for one hour. After drying under a stream of nitrogen, the derivatized sample was dissolved in toluene (20 uL) prior to GC-MS.

@@

The raw data and R code used for visualizations contained in this paper and R Markdown documents are available online at https://github.com/LiYingWang/GJBpottery to enable research reproducibility and transparency [@Marwick2017].

@@

```{r chromatograms-plots-four, echo=FALSE, fig.cap= "GC-MS chromatograms of fatty acids extracted from four potsherds from the Guijiabao and Gaoshan sites. "Cn:x" refers to fatty acid with n carbon atoms and x double bonds. "IS" indicates internal standard used for quantification, including FA C21:0 and 5-α Cholestane"}
knitr::include_graphics(here::here("analysis/figures/chromatograms_four.png"))
```

```{r chromatogram-specific, cache= TRUE}
# plot the specific one
chroma_specific <-
  meth_all %>%
  filter(name %in% c("SYG-TN13-E22-2#2")) %>% 
  ggplot(aes(Path, File)) +
  geom_line(size = 0.3) +
  ggrepel::geom_text_repel(data = highlight_SYG2_2, #ggrepel::geom_text_repel
            aes(label = FA),
            size = 3.5,
            nudge_y = 250000,
            show.legend = FALSE) +
  scale_y_continuous(labels = scales::comma_format(),
                     limits = c(0, 7500000),
                     breaks = seq(0, 7500000, 1000000)) +
  scale_x_continuous(limits = c(15, 50),
                     breaks = seq(15, 50, 5),
                     expand = c(0, 0.5)) + # don't log, many peaks and distort real counts
  labs(title = "SYG-TN13-E22-2#2", x = "retention time", y = "relative Intensity") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"))

@@

One pottery sample (SYG-TN13-E22-2#2) has much higher lipid concentrations (about `r FA_max_sum` μg/g) than the rest of the samples. A high yield of lipids might relate to the texture and type of ceramic used or better archaeological contexts for preservation. This sample also has a charred interior surface, indicative of a function associated with food processing. Long-chain fatty acids from C20:0 to C32:0 are present in that sample, with an even-over-odd carbon chain length predominance, but in lower amounts than in medium-chain components (Figure \@ref(fig:chromatograms-specific)). The existence of long-chain fatty acids with a high amount of C16:0 relative to C18:0 might be related to plant waxes [@Post1996; @Copley2005]. However, it could also be associated with degraded animal fat based on a strong predominance of medium-chain fatty acids [@Whelton2018; @Dudd1998]. Overall, the GC-MS results demonstrate strong evidence of degraded animal fats.

```{r chromatograms-specific, echo=FALSE, fig.cap= "The MS chromatogram of the potsherd from the Guijiabao site shows the presence of long-chain fatty acids. "Cn:x" refers to fatty acid with n carbon atoms and x double bonds. "IS" indicates internal standards used for quantification, including FA C21:0 and 5-α Cholestane"}
knitr::include_graphics(here::here("analysis/figures/chromatograms_specific.png"))
```

@@

# Discussion
<!--A brief summary of interesting results-->
The aim of this paper is to explore whether new crops, such as C4 plants (e.g., millets) or C3 plants (e.g., rice), or animals (e.g., pigs/horses) became an essential part of the local diet in Yanyuan during the Neolithic period. Our results show that food residues in potsherds from the sites at Guijiabao and Gaoshan were dominated by ruminant fats during the Neolithic period. Moreover, it is striking to discover that non-ruminant fats in a Guijiabao potsherd might have originated from C3 plants (see below). Although the lipid yields in most of our samples are generally low, interpretative amounts of yields from all of our samples support the assumption that these ceramics were used as cooking or storage vesssels. The low lipid yield could be attributed to the bottom of the vessels, where there were fewer residues than other parts of the vessel, for example, the rim [@Evershed2008experimental; @Kwak2020; @Charters1993]. But we also found that one sample from the vessel bottom dated to the Neolithic period contains a richer composition of long-chain fatty acids compared with those in other samples, reflecting better preservation or a much higher lipid yield. We recognize that there are limitations and challenges for determining lipid sources over time or differences across regions due to our small sample sizes and a lack of local modern references for animal fats and plant oils. In addition, a detailed analysis of faunal remains from the Guijiabao site is still in progress, which may constrain a comprehensive comparison between different types of evidence. Nevertheless, the general pattern of lipid profiles could still provide informative clues for a better understanding of culinary practices at the Guijiabao site.

<!--discussion about possible plant origins-->
While it has been determined that millets (C4 plants), including foxtail and broomcorn millets, were the dominant cereals, according to recent macro-botanical and phytolith studies at Guijiabao [@Guijiabao2016; @Hao2022; @Huan2022], we did not detect any lipid residues associated with C4 plants in our samples. Surprisingly, we identified non-ruminant fats from a Guijiabao potsherd that might have been from terrestrial plants/C3 plants, based on similar isotopic values reported by other studies in East Asia [@Shoda2018; @Kwak2020]. According to the results from flotation analysis at the Guijiabao site, it is possible that the source of C3 plants is rice or native wild weeds, such as pennisetum and chenopodiaceae. However, the seeds of wild weeds only account for a small part of the floral assemblage derived from flotation [@Guijiabao2016]. Despite stone knives and tools that were also found at the sites that might have been related to farming practices, more studies are needed to confirm their usage for harvesting crops [@Zhou2019; @Hao2022; @Lin2022]. Another possible source for non-ruminant fats could have been freshwater fish, a possibility which can be supported by the finding of fish bones and net sinkers at Guijiabao [@Guijiabao2017b]. But we did not identify any lipid biomarkers related to aquatic resources, such as long-chain $\omega$-(o-alkylphenyl) alkanoic acids or phytanic acid, from our potsherds [@Evershed2008marine; @Hansel2004; @Cramp2014; @Lucquin2016].

@@

We observed a mixture of ruminant fats and C3 plants, based on analyses of a number of Guijiabao potsherds that date to the historical period (ca. 1000 BP). It appears that C3 plants could have been processed more in the historical period than they could have been in the Neolithic period. In addition to rice, the source of C3 plants could also have included wheat, since it was introduced during the historical period [@JiangGaoshan2017]. A larger sample size is needed to determine whether there was a change in diet over time. The overall depleted δ^13^C values for potsherd samples associated with ruminant fats suggest that C3 plants might have been an important food source for local animals. Despite direct C4 plant signals not being found, analysis of one Guijiabao potsherd shows the source of C4 plant-fed ruminants during the Neolithic period. This indicates that either C3 or C4 plants could have been the food source for ruminants, but whether those animals were domesticated is unknown. The zooarchaeological analysis of faunal bones from Guijiabao is still in progress, the results of which which could provide more details about local animal species for comparison.

@@

# Acknowledgements
We thank the Chengdu Municipal Institute of Cultural Relics and Archaeology for their permission to access and analyze the pottery samples. We thank Julian Sachs in the Department of Oceanography at the University of Washington for his support in letting us use their laboratory for sample preparation and for conducting lipid extraction and analysis. We thank Matthew Wolhowe for his suggestions and assistance with the GC-based machines. We also want to express our gratitude to the anonymous reviewers for their comments that helped us improve this paper. This research is sponsored by the National Science and Technology Council (PE11006-0107).

>>>>

---
title: "Investigating Neolithic pottery use from Guijiabao in the Yanyuan Basin in southwest China using organic residue analysis"
author:
  - Li-Ying Wang:
      email: liyingw@asihp.net
      institute: [AcS]
      correspondence: true
  - Kuei-chen Lin:
      email: kueichenlin@asihp.net
      institute: [AcS]
      correspondence: false
  - Zhiqing Zhou:
      email: zhouzhiQ2000@163.com
      institute: [InsC]
      correspondence: false
institute:
  - AcS: Institute of History and Philology, Academia Sinica 
  - InsC: Chengdu Institute of Cultural Relics and Archaeology
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
    The spread of new crops or terrestial animals across areas in southwest China during the Neolithic period is one of the important archaeological issues to understand the process of adoption of new culinary practices in this region. It is widely accepted that rice was introduced into southwest China in 4700 BP at the latest and milletes came even earlier. However, to what extent new crops were used by local peoples at that time remains unclear. The excavation of Guijiabao in the Yanyuan basin in southwest China uncovered abundant materials with well-preserved stratigraphic records and has provided a unique opportunity to understand the local adoption of new crops. To further understand pottery use and food practices at Guijiabao, we investigated organic residues preserved in potsherds using lipid biomarker identification and compounds-specific stable isotope analysis. The results show signals of terrestrial animals in most potsherds and a C3 plant source in one potshard, during the Neolithic period. Our case study indicates that hunting might have been the primary means for obtaining subsistence, supplemented by rice/millet farming. This study also demonstrates an approach using organic residue analysis to explore pottery use and culinary practices during the Neolithic period in southwest China. A more systematic sampling of different types of potsherds and temporal phases with larger samples is needed to provide a clearer picture of ancient diets and subsistence.

@@

# Introduction
<!-- intro to the topic and research aim-->
Movements of people between geographically distinct areas are usually accompanied by the exchange of goods, technology, or ideas that could have cultural or economic impacts on local people groups involved in the interactions [@Burmeister2000]. The Hengduan Mountain ranges and valleys in southwestern China are geographical areas that connect the Qinghai-Tibetan Plateau and the Yunnan–Guizhou Plateau and by which movement to these plateaus must occur. They play an important role in the cultural exchanges and spread of crops into southwestern China and even further to mainland Southeast Asia [@Ma2021; @Lin2022; @Hao2022; @Tong1986]. Despite a long-standing debate on the routes of rice/millet dispersal from northwest China and the middle Yangtze valley to southwestern China [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming was practiced in southwest Sichuan around 4700 BP and in western Yunnan before 4600 BP [@Jiang2016; @YunnanMuseum1981; @Hao2022; @Zhang2010; @Dal2022; @Chen2015]. The Hengduan Mountain regions in the Neolithic period are therefore essential for understanding the processes of agricultural expansion. Recent archaeological evidence from the Yanyuan Basin in the middle Hengduan Mountains further reveals that millet and rice were present at roughly the same period as the grains found in Sichuan [@Huan2022; @Hao2022]. However, to what extent the two crops were incorporated into the local diet is still not fully understood.

To explore whether new crops or terrestrial animals became important food sources during the Neolithic period in the Yanyuan Basin, we investigated food residues within ceramic vessels from Guijiabao, a long-term occupation settlement with complete temporal sequences from the Neolithic period to the historical period in Yanyuan that have been excavated so far [@Guijiabao2017a; @Guijiabao2020a]. The new evidence at Guijiabao reveals that millet and rice were both present there in the Neolithic period around 5000 BP, earlier than previously thought [@Guijiabao2016], and millet could have already been part of the local diet [@Lin2022]. Although some of these analyses suggest the presence of mixed millet/rice farming [@Huan2022], we still lack direct evidence of culinary practices during this period. To understand the extent of consuming new crops or terrestrial animals at Guijiabao, we ask the questions: What were the culinary practices that can be deduced by the pottery? Can we identify the new crops or terrestrial animals that were incorporated into the local food diets during the Neolithic period? We answer these questions by examining potsherds using organic residue analysis, which allows us to directly explore food cooked or processed in ceramic vessels.

The recent growing application of organic residue analysis demonstrates its usefulness in understanding ancient pottery use [@Kimpe2004; @Mayyas2022], human subsistence strategies, or culinary practices across a region over time [@Kwak2020; @Craig2011; @Suryanarayan2021; @Zhang2022; @Junno2020]. Examining organic residues, particularly lipids, enables archaeologists to investigate ancient food practices through lipid distributions and specific lipid compounds as biomarkers for tracing food origins [@Copley2005; @Whelton2021]. This is possible due to the protection of clay matrices from microbiological degradation and water leaching, which allows lipid molecules to accumulate and be well preserved in the porous fabric of pottery for over a thousand years [@Evershed1993; @Evershed2008; @Roffet2017]. In this study, we detected and identified lipid residues absorbed into the ceramic matrix that can reveal the original contents cooked or stored in a vessel during its lifetime of use [@Dudd1999; @Miller2020]. Through our case study, we attempt to investigate culinary practices during a period of frequent cultural interactions that accompanied agricultural expansion.

@@

<!--Yanyuan area and Guijiabao site-->
We focus on the Guijiabao site located in the Yanyuan Basin, which is a pull-apart basin in the middle eastern part of the Hengduan Mountain ranges (Figure \@ref(fig:SW-china-map)). The Guijiabao site allows us to understand the early adoption of agriculture because of its uniqueness and rarity: it contains complete cultural components from the Neolithic period to the historical period in that region [@Guijiabao2017a; @Guijiabao2017b]. A survey in 2015 and subsequent excavations from 2016 to 2019 revealed abundant archaeological materials and features, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guijiabao was a settlement site [@Hao2022]. The chronology of the Guijiabao site can be divided into three phases, the Neolithic period (5000-3700 BP), the Bronze Age (3200-2700 BP), and the late North Song to early Ming dynasties (1200-500 BP) based on 19 radiocarbon dates from plant remains and a human bone. For the Neolithic period, it can be further divided into two subphases, 5000-4500 BP and 4500-3700 BP, according to the content of artifact assemblages and ceramic forms and styles that were identified [@Hao2022].

@@

Although the recent findings at the Guijiabao and Gaoshan sites support mixed agriculture as the primary subsistence strategy, it is still unclear how much millet or rice was included in the local diets during the Neolithic period. The macro-botanical and phytolith analyses of the Guijiabao site show that millet was the major crop, similar to the findings at the Yingpanshan site in Sichuan [@Zhao2011]. However, plant remains from many sites in southwest China reveal a predominance of rice, such as at the Gaoshan and Baodun sites in Sichuan [@JiangGaoshan2017] and Baiyangcun sites in Yunnan [@Jiang2016; @Henglanshan2014]. This indicates that there were variations in agricultural practices in southwest China. To better understand the extent of the adoption of new crops, organic residue analysis of ceramics is a useful approach since we can directly study food residues related to culinary practices, including food preparation, processing, and cooking. This provides another way to examine the expansion of agriculture in the Yanyuan area.

@@

For sample processing, the surface of each potsherd was removed to prevent external contamination from environmental factors and handling during excavation or post-excavation analysis. Approximately 3.5 grams of powdered samples were extracted with a mixture of dichloromethane/methanol solution (9:1 v/v) using accelerated solvent extraction (ASE Dionex 350; at 100 °C). We conducted saponification by adding 10 mL of 1M KOH in methanol and 1 mL water to the extracts and heating them in an oven at 70 °C overnight. Saponified extracts were acidified with 6M hydrochloric acid (HCl), and the acid fraction was extracted using hexane ($3\times10$ mL). A 5% aliquot of each sample was derivatized using 20 uL N,O-bis(trimethylsilyl)trifluoroacetamide (BSTFA) with 1% v/v trimethylchlorosilane (TMCS), and heated at 70 °C for one hour. After drying under a stream of nitrogen, the derivatized sample was dissolved in toluene prior to GC-MS.

@@

The raw data and R code used for visualizations contained in this paper and R Markdown documents are available online at https://github.com/LiYingWang/GJBpottery to enable research reproducibility and transparency [@Marwick2017; @MarwickandWang2022].

@@

```{r chromatograms-plots-four, echo=FALSE, fig.cap= "GC-MS chromatograms of fatty acids extracted from four potsherds from the Guijiabao and Gaoshan sites. 'Cn:x' refers to fatty acid with n carbon atoms and x double bonds. 'IS' indicates internal standard used for quantification, including FA C21:0 and 5-α Cholestane"}
knitr::include_graphics(here::here("analysis/figures/chromatograms_four.png"))
```

```{r chromatogram-specific, cache= TRUE}
# plot the specific one
chroma_specific <-
  meth_all %>%
  filter(name %in% c("SYG-TN13-E22-2#2")) %>% 
  ggplot(aes(Path, File)) +
  geom_line(size = 0.3) +
  ggrepel::geom_text_repel(data = highlight_SYG2_2, #ggrepel::geom_text_repel
            aes(label = FA),
            size = 3.5,
            nudge_y = 250000,
            show.legend = FALSE) +
  scale_y_continuous(labels = scales::comma_format(),
                     limits = c(0, 7500000),
                     breaks = seq(0, 7500000, 1000000)) +
  scale_x_continuous(limits = c(20, 50),
                     breaks = seq(20, 50, 5),
                     expand = c(0, 0.5)) + # don't log, many peaks and distort real counts
  labs(title = "SYG-TN13-E22-2#2", x = "retention time", y = "relative Intensity") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"))

@@

One pottery sample (SYG-TN13-E22-2#2) has much higher lipid concentrations (about `r FA_max_sum` μg/g) than the rest of the samples. A higher yield of lipids might relate to the texture and type of ceramic used or better archaeological contexts for preservation. This sample also has a charred interior surface, indicative of a function associated with food processing. Long-chain fatty acids from C20:0 to C32:0 are present in that sample, with an even-over-odd carbon chain length predominance, but in lower amounts than in medium-chain components (Figure \@ref(fig:chromatograms-specific)). The existence of long-chain fatty acids with a high amount of C16:0 relative to C18:0 might be related to plant waxes [@Post1996; @Copley2005]. However, it could also be associated with degraded animal fat based on a strong predominance of medium-chain fatty acids [@Whelton2018; @Dudd1998]. Overall, the GC-MS results demonstrate strong evidence of degraded animal fats.

```{r chromatograms-specific, echo=FALSE, fig.cap= "The MS chromatogram of the potsherd from the Guijiabao site shows the presence of long-chain fatty acids. 'Cn:x' refers to fatty acid with n carbon atoms and x double bonds. 'IS' indicates internal standards used for quantification, including FA C21:0 and 5-α Cholestane"}
knitr::include_graphics(here::here("analysis/figures/chromatograms_specific.png"))
```

@@

# Discussion
<!--A brief summary of interesting results-->
The aim of this paper is to explore whether new crops, such as C4 plants (e.g., millets) and C3 plants (e.g., rice), or animals (e.g., pigs/horses) became an essential part of the local diet in Yanyuan during the Neolithic period. Our results show that food residues in potsherds from the sites at Guijiabao and Gaoshan were dominated by ruminant fats during the Neolithic period. Moreover, it is striking to discover that non-ruminant fats in a Guijiabao potsherd might have originated from C3 plants (see below). Although the lipid yields in most of our samples are generally low, interpretative amounts of yields from all of our samples support the assumption that these ceramics were used as cooking or storage vesssels. The low lipid yield could be attributed to the bottom of the vessels, where there were fewer residues than other parts of the vessel, for example, the rim [@Evershed2008experimental; @Kwak2020; @Charters1993]. But we also found that one sample from the vessel bottom dated to the Neolithic period contains a richer composition of long-chain fatty acids compared with those in other samples, reflecting better preservation or a much higher lipid yield. We recognize that there are limitations and challenges for determining lipid sources over time or differences across regions due to our small sample sizes and a lack of local modern references for animal fats and plant oils. In addition, a detailed analysis of faunal remains from the Guijiabao site is still in progress, which may constrain a comprehensive comparison between different types of evidence. Nevertheless, the general pattern of lipid profiles could still provide informative clues for a better understanding of culinary practices at the Guijiabao site.

<!--discussion about possible plant origins-->
While it has been determined that millets (C4 plants), including foxtail and broomcorn millets, were the dominant cereals, according to recent macro-botanical and phytolith studies at Guijiabao [@Guijiabao2016; @Hao2022; @Huan2022], we did not detect any lipid residues associated with C4 plants in our samples. Surprisingly, we identified non-ruminant fats from a Guijiabao potsherd that might have been from terrestrial plants/C3 plants, based on similar isotopic values reported by other studies in East Asia [@Shoda2018; @Kwak2020]. According to the results from flotation analysis at the Guijiabao site, it is possible that the source of C3 plants is rice or native wild weeds, such as pennisetum and chenopodiaceae. However, the seeds of wild weeds only account for a small part of the floral assemblage derived from flotation [@Guijiabao2016]. Despite stone knives and tools that were also found at the sites that might have been related to farming practices, more studies are needed to confirm their usage for harvesting crops [@Zhou2019; @Hao2022; @Lin2022]. Another possible source for non-ruminant fats could have been freshwater fish, a possibility which can be supported by the finding of fish bones and net sinkers at Guijiabao [@Guijiabao2017b]. But we did not identify any lipid biomarkers related to aquatic resources, such as long-chain $\omega$-(o-alkylphenyl) alkanoic acids or phytanic acid [@Evershed2008marine; @Hansel2004; @Cramp2014; @Lucquin2016], from our potsherds.

@@

We observed a mixture of ruminant fats and C3 plants, based on analyses of a number of Guijiabao potsherds that date to the historical period (ca. 1000 BP). It appears that C3 plants could have been processed more in the historical period than they could have been in the Neolithic period. In addition to rice, the source of C3 plants could also have included wheat, since it was introduced into the region during the historical period [@JiangGaoshan2017]. A larger sample size is needed to determine whether there was a change in diet over time. The overall depleted δ^13^C values for potsherd samples associated with ruminant fats suggest that C3 plants might have been an important food source for local animals. Despite direct C4 plant signals not being found, analysis of one Guijiabao potsherd shows the source of C4 plant-fed ruminants during the Neolithic period. This indicates that either C3 or C4 plants could have been the food source for ruminants, but whether those animals were domesticated is unknown. The zooarchaeological analysis of faunal bones from Guijiabao is still in progress, the results of which could provide more details about local animal species for comparison.

@@

# Acknowledgements
We thank the Chengdu Municipal Institute of Cultural Relics and Archaeology for their permission to access and analyze the pottery samples. We thank Julian Sachs in the Department of Oceanography at the University of Washington for his support in letting us use their laboratory for sample preparation and for conducting lipid extraction and analysis. We thank Matthew Wolhowe for his suggestions and assistance with the GC-based machines. We also want to express our gratitude to the anonymous reviewers for their comments that helped us improve this paper. This research is sponsored by the National Science and Technology Council (MOST 109-2410-H-001-009-MY2).
